experDesign: helping performing experiments on batches

### 2021-04-14 paper.Rmd
Update content

---
title: "experDesign: helping performing experiments on collections"
output: 
  md_document:
    preserve_yaml: true
authors:
  - name: "Lluís Revilla Sancho"
    orcid: 0000-0001-9747-2570
    affiliation: '1, 2'
  - name: "Juan José Lozano Salvatella"
    orcid: 0000-0001-9747-2570
    affiliation: '1'
  - name: "Azucena Salas Martínez"
    orcid: 0000-0001-9747-2570
    affiliation: '2'
affiliations:
  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
    index: 1
  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
    index: 2
date: Sys.Date()
tags:
  - R
  - Batch effect
  - experiment design
bibliography: paper.bib
---

@@

The design of an experiment is critical for their success. However, after a correct design the collection of samples is a critical step where many errors and problems might affect the experiment result. Missing samples on experiments that must be carried out on batches might impact the results if not taken into account. experDesign helps to minimize batch effects for known variables assigning which samples go on which batch.

@@

**Blocking** groups samples that are equal according to one or more variables allowing to estimate the variation on the measurement from these variables. **Randomization** is a method to get the average effect by mixing the potential confounding variables. **Replication** increases the number of samples used on the experiment to better estimate the variation of the experiment. On some settings (clínical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.

@@

The package experDesign provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch. It is similar to the *anticluster* method for maximum variance. If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.

Adding replicates is a method to have a successful experiment. When the design is already done then the replicates are called technical replicates [@blainey2014]. Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement. If the technical replicates are distributed on several batches they allow to measure the batch effect. To select the technical replicates needed and from which samples the function `extreme_cases` is provided. For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.

experDesign also provides several small utilities to make it easier design the experiment in batches. For instance, it helps to calculate the samples to place on the minimum number of batches required with `sizes_batches`.

>>>>

---
title: "experDesign: helping performing experiments on batches"
output: 
  md_document:
    preserve_yaml: true
authors:
  - name: "Lluís Revilla Sancho"
    orcid: 0000-0001-9747-2570
    affiliation: '1, 2'
  - name: "Juan José Lozano Salvatella"
    orcid: 0000-0001-9747-2570
    affiliation: '1'
  - name: "Azucena Salas Martínez"
    orcid: 0000-0001-9747-2570
    affiliation: '2'
affiliations:
  - name: "Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas"
    index: 1
  - name: "Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS"
    index: 2
date: Sys.Date()
tags:
  - R
  - Batch effect
  - experiment design
bibliography: paper.bib
---

@@

The design of an experiment is critical for their success. However, after a correct design the collection of samples is a critical step where many errors and problems might affect the experiment result. Missing samples on experiments that must be carried out on batches might impact the results if not taken into account. *experDesign* helps to minimize batch effects for known variables assigning which samples go on which batch.

@@

**Blocking** groups samples that are equal according to one or more variables allowing to estimate the variation on the measurement from these variables. **Randomization** is a method to get the average effect by mixing the potential confounding variables. **Replication** increases the number of samples used on the experiment to better estimate the variation of the experiment. On some settings (clinical, agriculture, ...) several of these techniques are applied together to ensure the robustness of the study.

@@

The package *experDesign* provides the function `design` to distribute the samples on multiple batches so that each variable is homogeneous in each batch. It is similar to the *anticluster* method for maximum variance. If the experiment is carried out on a specific spatial distribution the `spatial` function distributes the samples homogeneously also by position similar to *Omixer*.

In addition to distribute the samples on batches *experDesign* provides with tools to add technical replicates. When the design is already done then the replicates are called technical replicates [@blainey2014]. Technical replicates are samples that are measured multiple times and allow to reduce the uncertainty of the measurement. If the technical replicates are distributed on several batches they allow to measure the batch effect and thus minimize it. To select the technical replicates needed and from which samples the function `extreme_cases` is provided. For easier usage the `replicates` designs an experiment with the number of replicates per batch desired.

*experDesign* also provides several small utilities to make it easier design the experiment in batches. For instance, it helps to calculate the samples to dictribute on the minimum number of batches required with `sizes_batches`.
